Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-09 paper/paper.tex
Section 3

\usepackage{amsmath, mathpartir}
%, amssymb, amsfonts, stmaryrd}
% \usepackage{mathpartir}
% \usepackage{bussproofs}

@@

\newcommand{\CatC}{\mathcal{C}}
\newcommand{\CatD}{\mathcal{D}}
\newcommand{\CatE}{\mathcal{E}}
\newcommand{\CatI}{\mathcal{I}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\iso}{\cong}
\newcommand{\ceiling}[1]{\lceil #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}
%\newcommand{\coq}[1]{\mintinline{coq}{#1}}

@@

\subsection{Dump}
Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given
by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of type
\coq{App F A ~> A}:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
The least fixed-point for a functor \coq{F} is an instance of an \coq{F}-algebra
where the structure map is an isomorphism. This is sometimes referred to as the
\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind
this name shortly. \footnote[1]{To be more precise, initial algebras are
isomorphisms by Lambek's lemma, but isomorphisms of functors do not necessarily
correspond to initial algebras.}  We define the least fixed-point of \coq{F} as
an inductive type:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
As an example, the initial \coq{F}-algebra for the
functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
operation \coq{Cons : A * LFix F -> LFix F}.

@@

A straightforward mechanisation of functors as defined in
Section~\ref{sec:rec-schemes} does not work in Coq due to the strict positivity
condition. In fact, if we simply require that functors are functions
\coq{F : Type -> Type}
satisfying the necessary identity and composition properties, then we will not
be able to construct least/greatest fixed points of this type, and we will not
be able to mechanise recursion schemes that follow this structure:
\begin{coqcode}
(* REJECTED due to negative occurrence of (LFix F) *)
Inductive LFix (F : Type -> Type) := LFix_in (lfix_out : F (LFix F)).
\end{coqcode}
One common workaround for this is the use of Mendler-style
recursion~\cite{Mendler91}. Roughly, the idea is to ``pack'' an abstract type
together with the functor that captures the structure of the recursion. From
this abstract type, we will be able to ``extract'' the contents of the functor.
This is, for example, the solution used by the PaCo library~\cite{PaCo}, and a
simplified version of it is the following:
\begin{coqcode}
Inductive MendlerLFix (F : Type -> Type) :=
  In Y (next : Y -> MendlerLFix F) (in_op : F Y).
\end{coqcode}
While this definition solves the strict positivity problem, it does not solve
the problems of: (1) reasoning about data/program equalities; and (2) extraction
to idiomatic OCaml code. To reason about program equalities, we would need to
reason about equalities of functions with \emph{some abstract} \coq{Y} as their
domain.  Furthermore, the extraction of \coq{F Y} will lead to unsafe casts in
the extracted OCaml code.

\paragraph{Containers} Another common solution is to the strict positivity
problem is the use \emph{containers}~\cite{AbbottAG05}.  A container is defined
by a type of \emph{shapes} and a \emph{family of position types} indexed by
shapes.
\begin{coqcode}
(Shape : Type) (Pos : Shape -> Type)
\end{coqcode}

For the functor \coq{F} we introduced in the previous paragraph we would
define \coq{Shape} as \coq{unit + A} indicating there are two constructors in
the data type and one of these contains a piece of data of type \coq{A}.
Moreover, we would define the positions as \coq{Pos (inl tt) = Empty_set}
and  \coq{Pos (inr a) = unit} indicating that the first constructor does not
have any type variables and the second constructor has one type variable.

At this point an \emph{extension} of this container is a functor defined as
follows:
\begin{coqcode}
Record App (X : Type) := MkApp { shape : Shape; contents : Pos shape -> X }.
\end{coqcode}
with the obvious action on morphisms given by post-composition
with \coq{contents}.
\begin{coqcode}
Definition fmap (f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{coqcode}
In our running example, the type \coq{App X}, for some \coq{X} has two
inhabitants.  The first is a pair composed by \coq{inl tt : Shape} and a
function \coq{Pos (inl tt) -> X}. This latter is in fact a function of
type \coq{Empty_set -> X}. This pair is, therefore isomorphic to the
type \coq{unit}, the type with only one inhabitant.  The second inhabitant is
the pair composed by \coq{inr a : Shape}, for some \coq{a : A}, and a
function \coq{Pos (inr a) -> X}.  This latter type is equal to \coq{unit -> X}
which represents the elements of \coq{X} and therefore is isomorphic to \coq{X}.
This particular pair of shapes and positions is  therefore isomorphic to the
type \coq{unit + A * X}.

The correspondence between containers and polynomial functors has been
formalised in this work and can be found in the accompanying code (file
Container.v). 
% 
%Generally, any functor that
%is isomorphic to a container extension is called a \emph{container functor}.
%Here we will call \emph{container functor} the functors defined by a specific
%container extension.

\subsection{Extractable Containers}
In the previous section we defined containers using dependent types.  However,
Ocaml's type system is not equipped to handle these. To solve this problem
Coq's code extraction mechanism will insert unsafe casts.

Consider instead representing the positions of a container using a decidable
validity predicate (\coq{valid}) which assigns shapes to positions. We use
boolean functions and coercions from \coq{bool} to \coq{Prop} to represent
decidable predicates, similarly to SSReflect~\cite{GonthierL09}. The family of
positions for a given shape in a container can be recovered by defining:
\begin{coqcode}
Class Cont (Shape : Type) (APos : Type) := { valid : Shape -> APos -> bool }.
Record Pos `(C : Cont Shape APos) (shape : Shape) :=
  ValidPos { val : APos; IsValid : valid shape pos }.
\end{coqcode}
Coq's code extraction will now be able to erase the validity predicate, and
generate OCaml code that is free of unsafe casts. The OCaml code extracted for
\coq{Pos} will now be exactly the code extracted for \coq{APos}.
The decidability of the validity predicate is crucial for our purposes of 
remaining axiom-free. To illustrate this, suppose that we need to show
the equality of two container extensions. We will need to show that, for
the same positions, they will produce the same result. In Coq, the goal
would look as follows: 
\begin{coqcode}
 k : Pos C s -> X
 P1, P2 : valid s p
 ---------------------------
 k (ValidPos p P1) = k (ValidPos p P2)
\end{coqcode}
If \coq{valid} was a regular proposition in \coq{Prop}, it would not possible
to prove the equality of \coq{P1} and \coq{P2}. However, by using a decidable
predicate, if we know that \coq{P1} and \coq{P2} are of type 
\coq{valid s p = true}, then we can prove without any axioms that 
\coq{P1 = P2 = eq_refl}.

% A container extension is then defined as follows:
% \begin{coqcode}
% Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
% Record App (X : Type) := { shape    : Shape;
%                           contents : {p | Valid shape p} -> X }.
% \end{coqcode}
% This approach solves the problem of the unsafe casts, since the code extracted
% for \coq{{p | Valid shape pos}} will be an OCaml singleton type defined as the
% OCaml equivalent to \coq{Pos}.

\subsubsection{Equality of Container Extensions}
Reasoning about the equality of container extensions is not entirely solved by
using decidable validity predicates to define the families of positions. In
general, we want to equate container extensions that have the same shape, and
that, for equal shape and position, they return the same element. To avoid the
use of the functional extensionality axiom, we capture this relation with the
following inductive proposition in Coq:
\begin{coqcode}
Inductive AppR `{F : Cont Sh P} {X} (x y : App F X) : Prop :=
| AppR_ext (Es : shape x = shape y)
           (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 = cont y e2).
\end{coqcode}
Note that  we do not care about the validity proof of the positions, only their
value. This is to simplify (slightly) our proofs. This relation is
trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.

However, the use of a different equality for container extensions now forces us
to deal with the fact that some types have different definitions of equality.
In particular, we want to reason about the equality of functions of types such
as \coq{App F A -> B} (or \coq{B -> App F A}). Since 
these types now come with their own equivalence, any function that manipulates
them needs to be \emph{respectful}. I.e.\ given 
\coq{R : X -> X -> Prop} and \coq{R' : Y -> Y -> Prop},
we want functions (morphisms) that satisfy the following property:
\begin{coqcode}
forall (x y : X), R x y -> R' (f x) (f y)
\end{coqcode}

@@

\subsubsection{Polynomial Types}
We define a number of equivalences for polynomial types.
%These equivalences are
%be used when building proper morphisms that manipulate their respective types.
\begin{coqcode}
Instance ext_eq (A : Type) `{eq_B : setoid B} : setoid (A -> B).
Instance pair_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A * B).
Instance sum_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A + B).
Instance prop_eq : setoid Prop.
Instance pred_sub `{eA : setoid A} {P : A -> Prop} : setoid {a : A | P a}.
\end{coqcode}
Most of the definitions that involve functions and polynomial types are
straightworward.  Identity and composition are defined as 
\coq{fun x => x} and \coq{fun f g x => f (g x)} 
respectively, and the proofs that they are proper morphisms is straightforward,
and automatically discharged by Coq. 
Products are built using function \coq{fun f g x => (f x, g x)},
with the projections being the standard Coq  \coq{fst} and
\coq{snd} functions. Similarly, sum injections are encoded using
Coq's \coq{inl} and \coq{inr} constructors, and
pattern matching on them uses the function 
\coq{fun f g x => match x with | inl y => f y | inr y => g y end}.
The proofs that these morphisms are proper are straightforward.  Finally, we
also provide functions for currying/uncurrying, and  flipping the arguments of
a proper morphism. We force most of our definitions to be inlined, to help
Coq's code extraction mechanism to inline as many of these combinators as
possible.

>>>>

\usepackage{amsmath, mathpartir, stmaryrd}
%, amssymb, amsfonts, stmaryrd}
% \usepackage{mathpartir}
% \usepackage{bussproofs}

@@

\newcommand{\CatC}{\mathcal{C}}
\newcommand{\CatD}{\mathcal{D}}
\newcommand{\CatE}{\mathcal{E}}
\newcommand{\CatI}{\mathcal{I}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\iso}{\cong}
\newcommand{\ceiling}[1]{\lceil #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}
%\newcommand{\coq}[1]{\mintinline{coq}{#1}}

@@

\subsection{Dump}
Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given
by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of type
\coq{App F A ~> A}:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
The least fixed-point for a functor \coq{F} is an instance of an \coq{F}-algebra
where the structure map is an isomorphism. This is sometimes referred to as the
\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind
this name shortly. \footnote[1]{To be more precise, initial algebras are
isomorphisms by Lambek's lemma, but isomorphisms of functors do not necessarily
correspond to initial algebras.}  We define the least fixed-point of \coq{F} as
an inductive type:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}

where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
As an example, the initial \coq{F}-algebra for the
functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
operation \coq{Cons : A * LFix F -> LFix F}.

@@

A straightforward mechanisation of functors as defined in
Section~\ref{sec:rec-schemes} does not work in Coq due to the strict positivity
condition. In fact, if we simply require that functors are functions
\coq{F : Type -> Type}
satisfying the necessary identity and composition properties, then we will not
be able to construct least/greatest fixed points of this type, and we will not
be able to mechanise recursion schemes that follow this structure:
\begin{coqcode}
(* REJECTED due to negative occurrence of (LFix F) *)
Inductive LFix (F : Type -> Type) := LFix_in (lfix_out : F (LFix F)).
\end{coqcode}

\noindent
One common workaround for this is the use of Mendler-style
recursion~\cite{Mendler91}. Roughly, the idea is to ``pack'' an abstract type
together with the functor that captures the structure of the recursion. From
this abstract type, we will be able to ``extract'' the contents of the functor.
This is, for example, the solution used by the PaCo library~\cite{PaCo}, and a
simplified version of it is the following:
\begin{coqcode}
Inductive MendlerLFix (F : Type -> Type) :=
  In Y (next : Y -> MendlerLFix F) (in_op : F Y).
\end{coqcode}

\noindent
While this definition solves the strict positivity problem, it does not solve
the problems of: (1) reasoning about data/program equalities; and (2) extraction
to idiomatic OCaml code. To reason about program equalities, we would need to
reason about equalities of functions with \emph{some abstract} \coq{Y} as their
domain.  Furthermore, the extraction of \coq{F Y} will lead to unsafe casts in
the extracted OCaml code.

\paragraph{Containers} Another common solution is to the strict positivity
problem is the use \emph{containers}~\cite{AbbottAG05}.  A container
$S \triangleright P$ is defined by a type of \emph{shapes} $S : \Type$ and a
\emph{family of position types} indexed by shapes $P : S \to \Type$.  An
\emph{extension} of this container is a functor
$\llbracket S \triangleright P \rrbracket$, whose action on objects is given by
$\llbracket S \triangleright P \rrbracket \; X = \Sigma s : S.\; P\;s \to X$,
and on morphisms is given by post-composition, keeping the shape the same
$\llbracket S \triangleright P \rrbracket \; f = \lambda (s,\;g).\;(s,\;f\circ g)$.

To explain the role of shapes and positions in containers, consider the functor
$F\; X = 1 + A \times X$.  An equivalent container, i.e.\ a container with an
extension that is isomorphic to $F$ requires a shape that can distinguish two
cases, indicating that there are two constructors to build an object of type
$F\;X$. For example, we could define $S_{F} = 1 + A$. We need to consider two
cases for the family of positions:  $P_{F}(\mathsf{inj}_{1}\;*) = 0$, because
there are no occurrences of $X$ in the left case; and
$P_{F}(\mathsf{inj}_{2}\;a) = 1$, because there is one occurrence of $X$ in the
right hand side. It is easy to see that
$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.

In general, every polynomial functor has an equivalent container, with an
isomorphic extension. We have mechanised this result, and this can  be found in
the accompanying code (file Container.v).
% 
%Generally, any functor that
%is isomorphic to a container extension is called a \emph{container functor}.
%Here we will call \emph{container functor} the functors defined by a specific
%container extension.

\subsection{Extractable Containers}
A straightforward encoding of containers in Coq would use dependent types  to
represent families of positions.  However, Ocaml's type system is not equipped
to handle these, which will lead to extracting OCaml code with unsafe casts.

Consider instead representing the positions of a container using a decidable
validity predicate which assigns shapes to positions. In Coq, we use
a boolean function and a coercion from \coq{bool} to \coq{Prop} to represent
the decidable predicate \coq{valid}, similarly to SSReflect~\cite{GonthierL09}. Given
a \coq{Shape : Type}, and a type of \emph{all possible positions} \coq{APos : Type},
the family of
positions for a given shape can be defined as a dependent pair:
\begin{coqcode}
Definition Pos (s : Shape) = { p : APos | valid s p }.
\end{coqcode}
Coq's code extraction will now be able to erase the validity predicate, and
generate OCaml code that is free of unsafe casts. The OCaml code extracted for
\coq{Pos} will now be exactly the code extracted for \coq{APos}.
The decidability of the validity predicate is crucial for our purposes of 
remaining axiom-free. To illustrate this, suppose that we need to show
the equality of two container extensions. We will need to show that, for
the same positions, they will produce the same result. In Coq, the goal
would look as follows: 
\begin{coqcode}
 k : Pos s -> X
 P1, P2 : valid s p
 ---------------------------
 k (existT _ p P1) = k (existT _ p P2)
\end{coqcode}
If \coq{valid} was a regular proposition in \coq{Prop}, it would not possible
to prove the equality of \coq{P1} and \coq{P2}. However, by using a decidable
predicate, if we know that \coq{P1} and \coq{P2} are of type 
\coq{valid s p = true}, then we can prove without any axioms that 
\coq{P1 = P2 = eq_refl}.

A container extension, \coq{App} in Coq, is then defined as follows:
\begin{coqcode}
Record App (X : Type)
   := { shape : Shape; contents : {p | valid shape p} -> X }.
% \end{coqcode}

This approach solves the problem of the unsafe casts, since the code extracted
for \coq{{p | valid shape p}} will be an OCaml singleton type defined as the
OCaml equivalent to \coq{Pos}.

\subsubsection{Equality of Container Extensions}
Reasoning about the equality of container extensions is not entirely solved by
using decidable validity predicates to define the families of positions. In
general, we want to equate container extensions that have the same shape, and
that, for equal shape and position, they return the same element. To avoid the
use of the functional extensionality axiom, we capture this relation with the
following inductive proposition in Coq:
\begin{coqcode}
Inductive AppR (x y : App F X) : Prop :=
| AppR_ext (shape x = shape y)
           (forall p1 p2, projT1 e1 = projT1 e2 -> contents p1 = contents p2).
\end{coqcode}
Note that  we do not care about the validity proof of the positions, only their
value. This is to simplify (slightly) our proofs. This relation is
trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.

However, the use of a different equality for container extensions now forces us
to deal with the fact that some types have different definitions of equality.
In particular, we want to reason about the equality of functions of types such
as \coq{App A -> B} (or \coq{B -> App A}). Since
these types now come with their own equivalence, any function that manipulates
them needs to be \emph{respectful}. I.e.\ given 
\coq{R : X -> X -> Prop} and \coq{R' : Y -> Y -> Prop},
we want functions (morphisms) that satisfy the following property:
\begin{coqcode}
forall (x y : X), R x y -> R' (f x) (f y)
\end{coqcode}

@@

\subsubsection{Polynomial Types}
We define a number of equivalences for polynomial types.
%These equivalences are
%be used when building proper morphisms that manipulate their respective types.
\begin{coqcode}
Instance ext_eq (A : Type) `{eq_B : setoid B} : setoid (A -> B).
Instance pair_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A * B).
Instance sum_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A + B).
Instance prop_eq : setoid Prop.
Instance pred_sub `{eA : setoid A} {P : A -> Prop} : setoid {a : A | P a}.
\end{coqcode}
Most of the definitions that involve functions and polynomial types are
straightworward.  Identity and composition are defined as 
\coq{fun x => x} and \coq{fun f g x => f (g x)} 
respectively, and the proofs that they are proper morphisms is straightforward,
and automatically discharged by Coq. 
Products are built using function \coq{fun f g x => (f x, g x)},
with the projections being the standard Coq  \coq{fst} and
\coq{snd} functions. Similarly, sum injections are encoded using
Coq's \coq{inl} and \coq{inr} constructors, and
pattern matching on them uses the function:
\begin{coqcode}
fun f g x => match x with | inl y => f y | inr y => g y end
\end{coqcode}
The proofs that these morphisms are proper are straightforward.  Finally, we
also provide functions for currying/uncurrying, and  flipping the arguments of
a proper morphism. We force most of our definitions to be inlined, to help
Coq's code extraction mechanism to inline as many of these combinators as
possible.
